Towards Automated Merging of Code Clones in Object-Oriented Programming Languages\\- Work in Progress -

### 2019-05-07 sattose/paper.tex
Ana+++

Code clones can be found anywhere in a codebase. The location of a clone in the code has an impact on how it can be refactored. Therefore, we first look at where and how often clones can be found, enabling the prioritization of refactoring opportunities.
% Using this information we can determine in which locations clones are found the most, on basis of which the refactoring will be prioritized. 
Furthermore, a duplicate fragment in a codebase does not always have to be an exact match with another fragment to be considered a clone. Therefore, we also analyse the impact of the different definitions of clone types on refactoring opportunities.
% We look at the definitions of different types of clones and the opportunities to refactor duplicate parts of code albeit not an exact match.

@@

A higher type of clone means that it is harder to detect. It also makes the clone harder to refactor, as more transformations would be required. Higher clone types also become more disputable whether they actually indicate a harmful antipattern (as not every clone is harmful \cite{jarzabek2010clones, kapser2008cloning}).
\subsection{Related work in clone refactoring tools}
The Duplicated Code Refactoring Advisor (DCRA) looks into refactoring opportunities for clone pairs \cite{fontana2012duplicated}. This tool only focuses on refactoring clone pairs, the authors arguing that \textit{clone pairs are much easier to manage when considered singularly.} As intermediate steps, the authors measure a corpus of Java systems for some clone-related properties of the systems, like the relation (in terms of inheritance) between code fragments in a clone pair.

Aries \cite{higo2004aries, higo2008metric} focuses on the detection of refactoring oriented clones. On basis of the relation between clone instances through inheritance, similar to Fontana et al., Aries proposes a refactoring. For instance, if two clone classes are siblings of each other (share the same superclass), they propose to perform ``Extract method'' and ``Pull up method'' sequentially. This tool only proposes such a refactoring, and does not provide help in the process of applying the refactoring.

@@

\section{Clone Detection}
As duplication in source code is a serious problem in many software systems, there are a lot of researches that look into code clones. Many tools have been proposed to detect various types of code clones. Two surveys of modern clone detection tools \cite{sheneamer2016survey, svajlenko2014evaluating} together show an overview of the most-popular clone detection tools up until 2016.

@@

@@

\footnotetext[3]{This clone detection tool has also been used by Fontana et al. \cite{fontana2012duplicated}}

\footnotetext{\label{note1}CPD is a Clone Detection tool that is part of PMD. PMD is a widely used tool for static code analysis in Java.}

@@

We also want to argue that statements that are not common between two clone instances, should not count towards the size of the clone (and thus towards the threshold which determines whether the clone will be taken into account). Also, clones should not start and not end with an added statement (as that would be nonsense: such a thing could be done for any clone).

@@

\section{Code Clone Context}
To be able to refactor code clones, it is very important to consider the context of the clone. We define the following aspects of the clone as its context:
\begin{enumerate}
  \item The relation of clone instances among each other through inheritance (for example: a clone instance resides in a superclass of another clone instance in the same clone class).
  \item Where a clone instance occurs in the code (for example: a method-level clone is a clone instance that is in a method).
  \item The contents of a clone instance (for example: the clone instance spans several methods).
\end{enumerate}
Figure \ref{fig:clonecontext} shows an abstract representation of clone classes and clone instances. The relation of clones through inheritance is measured on clone class level: it involves all child clone instances. The location and contents of clones is measured on clone instance level. A clone's location involves the file it resides in and the range it spans (for example: line 6 col 2 - line 7 col 50). A clone instance contents consists of a list of all statements and declarations it spans.
\begin{figure}[H]
    \includegraphics[width=1\columnwidth]{img/context}
  \caption{Abstract representation of clone classes and clone instances.}
  \label{fig:clonecontext}
\end{figure}
Everything in the context of a clone has a big impact on how it has to be refactored. For this study, we performed measurements on the context of clones in a large corpus of open source projects.

@@

\subsection{Relations Between Clone Instances}
When merging code clones in object-oriented languages, it is very important to consider the relation between clone instances. This relation has a big impact on how a clone should be merged, in order to improve the software design in the process.

\subsubsection{Categorizing Clone Instance Relations}
Fontana et al. \cite{fontana2015duplicated} describe measurements on 50 open source projects on the relation of clone instances to each other. To do this, they first define several categories for the relation between clone instances in object-oriented languages. A few of these categories are shown in figure \ref{fig:clonerelation}. These categories are as follows:
\begin{enumerate}
  \item \textbf{Same method}: All instances of the clone class are in the same method.
  \item \textbf{Same class}: All instances of the clone class are in the same class.
  \item \textbf{Superclass}: All instances of the clone class are children and parents of each other.
  \item \textbf{Ancestor class}: All instances of the clone class are superclasses except for the direct superclass.
  \item \textbf{Sibling class}: All instances of the clone class have the same parent class.
  \item \textbf{First cousin class}: All instances of the clone class have the same grandparent class.
\item \textbf{Common hierarchy class}: All instances of the clone class belong to the same hierarchy, but do not belong to any of the other categories.
\item \textbf{Same external superclass}: All instances of the clone class have the same superclass, but this superclass is not included in the project but part of a library.
\item \textbf{Unrelated class}: There is at least one instance in the clone class that is not in the same hierarchy.
\end{enumerate}

@@

\subsubsection{Our measurements}
We use a similar setup to that used by Fontana et al. (Table~3 of Fontana et al. \cite{fontana2015duplicated}). Fontana et al. measure clones using their own tool (DCRA). As explained in section~\ref{ch:tool-overview}, we chose to implement our own tool, CloneRefactor. Therefore, the setup for our measurements differs as follows from Fontana et al.:
\begin{itemize}
  \item We consider clone classes rather than clone pairs. The rationale for this is given in section \ref{chap:cloneclasses}.
\item We use different thresholds regarding when a clone should be considered. Fontana et al. seek clones that span a minimum of 7 source lines of code (SLOC). We seek clones with a minimum size of 6 statements/declarations. This is further explained in section \ref{chap:thresholds}.
\item We seek duplicates by statement/declaration rather than SLOC. This makes our analysis depend less on the coding style (in terms of newline usage) of the author of the software project.
\item We test a broader range of projects. Fontana et al. use a set of 50 relatively large projects. We use the corpus as explained in \ref{chap:corpus}, which contains a diverse set of projects (diverse both in volume and code quality).
\end{itemize}

Table~\ref{table:relations} contains our results regarding the relations between clone instances.

@@

\subsection{Clone instance location}
After mapping the relations between individual clones, we looked at the location of individual clone instances. A paper by Lozano et al. \cite{lozano2007evaluating} discusses the harmfulness of cloning. The authors argue that 98\% are produced at method-level. However, this claim is based on a small dataset and based on human copy-paste behavior rather than static code analysis. We validated this claim over the corpus we use. For this, we chose the following categories:
\begin{enumerate}
  \item \textbf{Method/Constructor Level:} A clone instance that does not exceed the boundaries of a single method or constructor (optionally including the declaration of the method or constructor itself).
  \item \textbf{Class Level:} A clone instance in a class, that exceeds the boundaries of a single method or contains something else in the class (like field declarations, other methods, etc.).
  \item \textbf{Interface Level:} A clone that is (a part of) an interface.
  \item \textbf{Enumeration Level:} A clone that is (a part of) an enumeration.
\end{enumerate}

The results for the clone instance locations are shown in table \ref{table:locations}.

@@

\subsection{Clone instance contents}
Finally, we looked at the contents of individual clone instances: what kind of declarations and statements do they span. We selected the following categories to be relevant for refactoring:
\begin{enumerate}
  \item \textbf{Full Method/Class/Interface/Enumeration:} A clone that spans a full class, method, constructor, interface or enumeration, including its declaration.
  \item \textbf{Partial Method/Constructor:} A clone that spans a method partially, optionally including its declaration.
  \item \textbf{Several Methods:} A clone that spans over two or more methods, either fully or partially, but does not span anything but methods (so not fields or anything in between).
  \item \textbf{Only Fields:} A clone that spans only global variables.
  \item \textbf{Includes Fields/Constructor:} A clone that spans a combination of fields and other things, like methods.
  \item \textbf{Method/Class/Interface/Enumeration Declaration:} A clone that contains the declaration (usually the first line) of a class, method, interface or enumeration.
  \item \textbf{Other:} Anything that does not match with above-stated categories.
\end{enumerate}

@@

However, not always can a clone easily be merged. Sometimes a clone spans a statement partially (like a for-loop of which only it's declaration and a part of the body is cloned). Merging the clones can be harder in such instances. Also, the cloned code can contain statements like labels, return, break, continue, etc. In such instances, more conditions may apply to be able to conduct a refactoring, if advisable at all.

>>>>

Code clones can be found anywhere in a codebase. The location of a clone in the code has an impact on how it can be refactored. Therefore, we first look at where and how often clones can be found, enabling the prioritization of refactoring opportunities.
% Using this information we can determine in which locations clones are found the most, on basis of which the refactoring will be prioritized. 
Furthermore, a duplicate fragment in a codebase does not always have to be an exact match with another fragment to be considered a clone. Therefore, we also analyze the impact of the different definitions of clone types on refactoring opportunities.
% We look at the definitions of different types of clones and the opportunities to refactor duplicate parts of code albeit not an exact match.

@@

A higher type of clone means that it is harder to detect. It also makes the clone harder to refactor, as more transformations would be required. Higher clone types also become more disputable whether they actually indicate a harmful antipattern (as not every clone is harmful \cite{jarzabek2010clones, kapser2008cloning}).
\subsection{Related work in clone refactoring tools}
The Duplicated Code Refactoring Advisor (DCRA) looks into refactoring opportunities for clone pairs \cite{fontana2012duplicated, fontana2015duplicated}. This tool only focuses on refactoring clone pairs, the authors arguing that \textit{clone pairs are much easier to manage when considered singularly.} As intermediate steps, the authors measure a corpus of Java systems for some clone-related properties of the systems, like the relation (in terms of inheritance) between code fragments in a clone pair.

Aries \cite{higo2004aries, higo2008metric} focuses on the detection of refactoring oriented clones. On basis of the relation between clone instances through inheritance, similar to Fontana et al. \cite{fontana2012duplicated}, Aries proposes a refactoring. For instance, if two clone classes are siblings of each other (share the same superclass), they propose to perform ``Extract method'' and ``Pull up method'' sequentially. This tool only proposes such a refactoring, and does not provide help in the process of applying the refactoring.

@@

\section{Clone Detection}
As duplication in source code is a serious problem in many software systems, many research efforts look into code clones. Many tools have been proposed to detect various types of code clones. Two surveys of modern clone detection tools \cite{sheneamer2016survey, svajlenko2014evaluating} together show an overview of the most-popular clone detection tools up until 2016. \todo{what is insufficient about these two? what triggered the need to reassess clone detectors? where are the four criteria coming from?}

@@

\footnotetext[3]{This clone detection tool has also been used by Fontana et al. \cite{fontana2012duplicated}}
\footnotetext{\label{note1}CPD is a Clone Detection tool that is part of PMD. PMD is a widely used tool for static code analysis in Java.}

@@

@@

We also argue that statements that are not common between two clone instances, should not count towards the size of the clone (and thus towards the threshold which determines whether the clone will be taken into account). Also, clones should not start and not end with an added statement (as that would be nonsense: such a thing could be done for any clone).

@@

\section{Relation, Location and Content Analysis of Clones}\label{chap:clonecontextexpl}
To be able to refactor code clones, it is very important to consider the context of the clone. We define the following aspects of the clone as its context:
\begin{enumerate}
  \item The relation of clone instances among each other through inheritance (for example: a clone instance resides in a superclass of another clone instance in the same clone class).
  \item Where a clone instance occurs in the code (for example: a method-level clone is a clone instance that is in a method).
  \item The contents of a clone instance (for example: the clone instance spans several methods).
\end{enumerate}
Figure \ref{fig:clonecontext} shows an abstract representation of clone classes and clone instances. The relation of clones through inheritance is measured on clone class level: it involves all child clone instances. The location and contents of clones is measured on clone instance level. A clone's location involves the file it resides in and the range it spans (for example: line 6 col 2 - line 7 col 50). A clone instance contents consists of a list of all statements and declarations it spans.
\begin{figure}[H]
    \includegraphics[width=1\columnwidth]{img/context}
  \caption{Abstract representation of clone classes and clone instances.}
  \label{fig:clonecontext}
\end{figure}
%\todo{do we have proof of this statement? if not, say that this is our intuition and a first step to prove it is to analyze a large body of projects} Everything in the context of a clone has a big impact on how it has to be refactored. 
% I guess it's a bit of an empty statement. Of course where some code is located has an impact on how to refactor it. Maybe remove it?
We analyzed the context of clones in a large corpus of open source projects. For these experiments, we used the CloneRefactor tool. These experiments map the context, as follows: the relation between clone instances is explained, measured and discussed in section\ref{chap:relationsinstances}; the location of clone instances is explained, measured and discussed in section \ref{chap:clonelocation};  the content of clone instances are explained, measured and discussed in section \ref{chap:clonecontents}.

@@

\subsection{Relations Between Clone Instances} \label{chap:relationsinstances}
When merging code clones in object-oriented languages, it is very important to consider the relation between clone instances. This relation has a big impact on how a clone should be merged, in order to improve the software design in the process. In this section, we display measurements we conducted on the corpus introduced in section \ref{chap:corpus}. These measurements are based on an experiment by Fontana et al. \cite{fontana2015duplicated}, which we will briefly introduce in section \ref{chap:catcloneinstancerelations}. We use a vastly different setup, which is explained in section \ref{chap:oursetup}. We then show our results in section \ref{chap:ourmeasurements}.

\subsubsection{Categorizing Clone Instance Relations}\label{chap:catcloneinstancerelations}
Fontana et al. \cite{fontana2015duplicated} describe measurements on 50 open source projects on the relation of clone instances to each other. To do this, they first define several categories for the relation between clone instances in object-oriented languages. A few of these categories are shown in figure \ref{fig:clonerelation}. These categories are as follows:
\begin{enumerate}
  \item \textbf{Same method}: All instances of the clone class are in the same method.
  \item \textbf{Same class}: All instances of the clone class are in the same class.
  \item \textbf{Superclass}: All instances of the clone class are children and parents of each other.
  \item \textbf{Ancestor class}: All instances of the clone class are superclasses except for the direct superclass.
  \item \textbf{Sibling class}: All instances of the clone class have the same parent class.
  \item \textbf{First cousin class}: All instances of the clone class have the same grandparent class.
\item \textbf{Common hierarchy class}: All instances of the clone class belong to the same hierarchy, but do not belong to any of the other categories.
\item \textbf{Same external superclass}: All instances of the clone class have the same superclass, but this superclass is not included in the project but part of a library.
\item \textbf{Unrelated class}: There is at least one instance in the clone class that is not in the same hierarchy.
\end{enumerate}

@@

\subsubsection{Our setup}\label{chap:oursetup}
We use a similar setup to that used by Fontana et al. (Table~3 of Fontana et al. \cite{fontana2015duplicated}). Fontana et al. measure clones using their own tool (DCRA). As explained in section~\ref{ch:tool-overview}, we chose to implement our own tool, CloneRefactor. Therefore, the setup for our measurements differs as follows from Fontana et al.:
\begin{itemize}
  \item We consider clone classes rather than clone pairs. The rationale for this is given in section \ref{chap:cloneclasses}.
\item We use different thresholds regarding when a clone should be considered. Fontana et al. seek clones that span a minimum of 7 source lines of code (SLOC). We seek clones with a minimum size of 6 statements/declarations. This is explained detail in section \ref{chap:thresholds}.
\item We seek duplicates by statement/declaration rather than SLOC. This makes our analysis depend less on the coding style (in terms of newline usage) of the author of the software project.
\item We test a broader range of projects. Fontana et al. use a set of 50 relatively large projects. We use the corpus as explained in \ref{chap:corpus}, which contains a diverse set of projects (diverse both in volume and code quality).
\end{itemize}

\subsubsection{Our results} \label{chap:ourmeasurements}
Table~\ref{table:relations} contains our results regarding the relations between clone instances.

@@

\subsection{Clone instance location}\label{chap:clonelocation}
After mapping the relations between individual clones, we looked at the location of individual clone instances. A paper by Lozano et al. \cite{lozano2007evaluating} discusses the harmfulness of cloning. The authors argue that 98\% are produced at method-level. However, this claim is based on a small dataset and based on human copy-paste behavior rather than static code analysis. We validated this claim over our corpus. The results for the clone instance locations are shown in table \ref{table:locations}. We chose the following categories:
\begin{enumerate}
  \item \textbf{Method/Constructor Level:} A clone instance that does not exceed the boundaries of a single method or constructor (optionally including the declaration of the method or constructor itself).
  \item \textbf{Class Level:} A clone instance in a class, that exceeds the boundaries of a single method or contains something else in the class (like field declarations, other methods, etc.).
  \item \textbf{Interface Level:} A clone that is (a part of) an interface.
  \item \textbf{Enumeration Level:} A clone that is (a part of) an enumeration.
\end{enumerate}

@@

\subsection{Clone instance contents}\label{chap:clonecontents}
Finally, we looked at the contents of individual clone instances: what kind of declarations and statements do they span. We selected the following categories to be relevant for refactoring:
\begin{enumerate}
  \item \textbf{Full Method/Class/Interface/Enumeration:} A clone that spans a full class, method, constructor, interface or enumeration, including its declaration.
  \item \textbf{Partial Method/Constructor:} A clone that spans a method partially, optionally including its declaration.
  \item \textbf{Several Methods:} A clone that spans over two or more methods, either fully or partially, but does not span anything but methods (so not fields or anything in between).
  \item \textbf{Only Fields:} A clone that spans only global variables.
  \item \textbf{Includes Fields/Constructor:} A clone that spans a combination of fields and other things, like methods.
  \item \textbf{Method/Class/Interface/Enumeration Declaration:} A clone that contains the declaration (usually the first line) of a class, method, interface or enumeration.
  \item \textbf{Other:} Anything that does not match with above-stated categories.
\end{enumerate}

@@

However, a clone is not always easily merged. Sometimes a clone spans a statement partially (like a for-loop of which only it's declaration and a part of the body is cloned). Merging the clones can be harder in such instances. Also, the cloned code can contain statements like labels, return, break, continue, etc. In such instances, more conditions may apply to be able to conduct a refactoring, if advisable at all.
